---
title: "Final Project - Predicting Win Rates in La Liga"
author: "Osarodion"
date: "4/18/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries}
# install.packages("ggrepel")
library(ggplot2)
library(dplyr)
library(corrplot)
library(skimr)
library(MLmetrics)
library(ggrepel)

```



Advanced Business Analytics – Final Project
===========================================

Team Members
------------
Osarodion Ogbebor Evans

Project Title
-------------
Predicting win rates for the 2017/2018 Spanish la liga season

Type of Final Project
---------------------
Type 1 - Analysis of a data set of interest

Executive Summary of the Proposed Project
-----------------------------------------

The Spanish la liga is often referred to as the most competitive football (soccer) league in the world. Home to the most successful football teams in Europe, the league is often considered the toughest. Winning the league requires acquiring the most points from all games played in the season. For every season, there is a total of 38 games for every team .i.e. 19 home games and 19 away games. For every team, a win earns 3 points, a draw earns 1 point and a loss earns 0 point.

With lesser goals scored and more goals conceded than the runner-up, Real Madrid was the crowned champion of the 2016-2017 la liga Season. 2017-2018 season was won by Barcelona with more goals scored and more goals conceded than the runner-up (Atletico Madrid). However, Real Madrid ended the season in 3rd place having scored 30 goals more than the runner-up, and 5 goals less than the season champion.

This project analyzes different match statistics and their effects on the number of points earned in a season. It covers the use of R in creating different predictive models, and the use of these models in predicting season points for different seasons.

Data Source
-----------------------
The data required for this project is a table consisting of different statistics for every game of the 2016/2017, 2017/2018 & 2019/2019 la liga primera division seasons. This data is extracted from [footystats](https://footystats.org/download-stats-csv)

```{r read_csv}
#read csv files
laliga1617 <- read.csv("Data/spain-la-liga-teams-2016-to-2017-stats.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE)
laliga1718 <- read.csv("Data/spain-la-liga-teams-2017-to-2018-stats.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE)
laliga1819 <- read.csv("Data/spain-la-liga-teams-2018-to-2019-stats.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE)

```

Data  Description and Selection
-------------------------------
For this project, the 'laliga1617' and 'laliga1718' will be used as the train and test data sets respectively. 

```{r data_structure}
#Display structure, top and bottom rows of both train & test data sets
str(laliga1617)
str(laliga1718)
 
# head(laliga1617)
# head(laliga1718)

# tail(laliga1617)
# tail(laliga1718)

```

The variables (predictors) in the csv files represent game statistics and betting odds recorded for the season. The analysis and models for this project will be based only on the game statistics. Among the game statistics, most variables have components such as home and away values. For example columns 9 through 11 show wins, home wins and away wins, where wins = home wins + away wins.

```{r display_wins}
head(laliga1617[,9:11])
```

I will be reassigning only the whole variables for the game statistics to the laliga1617 data frame (i.e. excluding the betting odds, and the home and away variables for most of the game statistics). In part, this is to reduce the effect of multi-collinearity, which will be addressed in another section.

```{r assign_game_stats}
laliga1617 <- laliga1617 %>% select(1,9,12,15,18,21,25,26,27,37,40,43,55,58,61,64,67,70,73,76,79,82,85,88,91)
head(laliga1617)
str(laliga1617)

```

The response variable for the predictive models will be season points. As described in the executive summary, a draw wins both teams 1 point, and a win earns the winning team 3 points. Therefore season points can be calculated as a sum of (3 x the number of games won) and (the number of games drawn).

```{r win_rate}
#calculating win rate for each team.
laliga1617 <- laliga1617 %>%
  mutate(season_points = (3*wins)+draws)
laliga1718 <- laliga1718 %>%
  mutate(season_points = (3*wins)+draws)
laliga1819 <- laliga1819 %>%
  mutate(season_points = (3*wins)+draws)

```

Exploratory Data Analysis (EDA)
--------------------------
The EDA provides a summary of our data, and an investigation into the relationships between the predictors and our response variable.

```{r summary_stats}
summary(laliga1617)

```

An observation from the summary is the absence of missing data, which can affect statistical power and lead to bias in estimating coefficients.

```{r correlation_plot}
cormat <- cor(laliga1617[,2:26])
# cormat
corrplot::corrplot(cormat, )

```

The correlation plot provides insights on the effect of other predictors (excluding wins, draws and losses) on win rate. Scatter plots are provided for further exploration.

```{r barplots}
g <- ggplot(laliga1617)
g + geom_bar(aes(reorder(team_name,season_points), y=season_points), stat = 'identity') + xlab("team_name") + coord_flip()
g + geom_bar(aes(reorder(team_name,wins), y=wins), stat = 'identity') + xlab("team_name") + coord_flip()
g + geom_bar(aes(reorder(team_name,losses), y=losses), stat = 'identity') + xlab("team_name") + coord_flip()
g + geom_bar(aes(reorder(team_name,draws), y=draws), stat = 'identity') + xlab("team_name") + coord_flip()
g + geom_bar(aes(reorder(team_name,goals_scored), y=goals_scored), stat = 'identity') + xlab("team_name") + coord_flip()

```

Shot analysis

```{r barplots_by_shots}
g + geom_bar(aes(reorder(team_name,shots), y=shots,fill=shots_on_target), stat = 'identity') + xlab("team_name") + coord_flip()
g + geom_bar(aes(reorder(team_name,shots), y=shots,fill=shots_off_target), stat = 'identity') + xlab("team_name") + coord_flip()
g + aes(x=points_per_game, y=shots_on_target)+geom_point()+geom_smooth(method = "lm")+geom_text_repel(aes(label=team_name), size=3)
g + aes(x=points_per_game, y=shots_off_target)+geom_point()+geom_smooth(method = "lm")+geom_text_repel(aes(label=team_name), size=3)

```

Predictor Analysis

```{r scatter_plots}
g + aes(x=goals_scored, y=season_points)+geom_point()+geom_smooth(method = "lm")
g + aes(x=goals_conceded, y=season_points)+geom_point()+geom_smooth(method = "lm")
g + aes(x=clean_sheets, y=season_points)+geom_point()+geom_smooth(method = "lm")
g + aes(x=corners_total, y=season_points)+geom_point()+geom_smooth(method = "lm")
g + aes(x=average_possession, y=season_points)+geom_point()+geom_smooth(method = "lm")
g + aes(x=shots_on_target, y=season_points)+geom_point()+geom_smooth(method = "lm")
g + aes(x=shots_off_target, y=season_points)+geom_point()+geom_smooth(method = "lm")
g + aes(x=goal_difference_half_time, y=season_points)+geom_point()+geom_smooth(method = "lm")
g + aes(x=leading_at_half_time, y=season_points)+geom_point()+geom_smooth(method = "lm")
g + aes(x=cards_total, y=season_points)+geom_point()+geom_smooth(method = "lm")
g + aes(x=shots, y=season_points)+geom_point()+geom_smooth(method = "lm")

```

The shape of the confidence interval has been included to the scatter plots to aid with comprehending variations. Contrary to the notion of Spanish football, where high ball possession is perceived as a principal contributor to winning games, it was interesting to see average possession and total corner kicks with thick bands, indicating both predictors have lower correlation with season points than other plotted predictors. (N.B. While correlation analysis can provide a summary of the relationship between predictors, it can only determine direct relationship. In regression, we are seeking to determine how much the dependent variable changes given a change in the independent variable. i.e. "y" as a function of "x")

Model building & assessment
---------------------------
Two models would be considered in building the predictive models - the base model(null model) and the full model.

```{r models}
nullModel <- lm(season_points ~ 1, data = laliga1617)
fullModel <- lm(season_points ~ goals_scored + goals_conceded + clean_sheets + corners_total + average_possession+
                shots_on_target+shots_off_target+goal_difference_half_time+leading_at_half_time+cards_total,
                data = laliga1617)

```

```{r model_summary}
summary(nullModel)
summary(fullModel)
```

The nullModel provides a base comparison for the full model. It indicates an intercept of 53 season points. The fullModel has a high R-squared value and a low p-value which indicate the model fits well and is a good predictor of season points. The value of the coefficient of average_possession qualifies it as a good predictor as compared to its performance in the correlation analysis. A key observation from the summary of the fullModel is multi-collinearity. This would be addressed in another section.

```{r residual_analysis}
plot(nullModel$residuals, col="blue")
plot(fullModel$residuals, col="green")

```

The randomness presented by the residual plots support the R-squared value, and indicates the model fits.

```{r MAE_values}
MAE_models <- c(MAE(laliga1617$season_points, nullModel$fitted.values),
                MAE(laliga1617$season_points, fullModel$fitted.values))

MAE_result <- data.frame(MAE_models[1],MAE_models[2])
names(MAE_result) <- c("nullModel","fullModel")
MAE_result
```
The value of the Mean Absolute Error (MAE) for the fullModel is significantly lower than that of the nullModel. This validates the fullModel as a better predictor.

Assessing Multi-Collinearity
------------------------------
```{r summary_assessingMcol}
corrplot(cormat)
summary(fullModel)
```

Some predictors (such as "shots_on_target", "shots_off_target", "leading_at_half_time") with a positive linear association with season points, as seen in the correlation plot, have been estimated to have a negative linear association with season points, as presented by the direction of their estimated values (-ve) in the summary result. This is as a result of multi-collinearity. "shots_on_target" & "shots_off_target" are highly correlated with themselves. While "leading_at_half_time" is highly correlated with "goal_difference_half_time". To resolve this, a new model would be created with "shots" replacing "shots_on_target" & "shots_off_target", and without "leading_at_half_time".

```{r newModel}
newModel <- lm(season_points ~ goals_scored + goals_conceded + clean_sheets + corners_total + average_possession+
                shots+goal_difference_half_time+cards_total,
                data = laliga1617)
summary(newModel)
```

The MAE value of the newModel is compared with the nullModel and fullModel.

```{r MAE_newModel}
MAE_result$newModel <- MAE(laliga1617$season_points, newModel$fitted.values)

MAE_result
```

The difference in the MAE values of the fullModel and newModel can be considered insignificant. With this result, the newModel will be used in predicting season points for the 2017/2018 & 2018/2019 la liga seasons

Predicting Season Points
-------------------------

```{r predict}
predict_laliga1718 <- predict(newModel, newdata = laliga1718)
predict_laliga1819 <- predict(newModel, newdata = laliga1819)

```


```{r MAE_predicted}
MAE_laliga1718 <- MAE(laliga1718$season_points, predict_laliga1718)
MAE_laliga1819 <- MAE(laliga1819$season_points, predict_laliga1819)
sprintf("MAE value of predicted 2017/2018 season points to actual season points: %.2f", MAE_laliga1718)
sprintf("MAE value of predicted 2018/2019 season points to actual season points: %.2f", MAE_laliga1819)
```

```{r act_vs_pred_season_points}
act_vs_pred_1718 <- data.frame(Team_name = laliga1718[,"team_name"],
                               actual_season_points = laliga1718[,"season_points"],
                               predicted_season_points = predict_laliga1718)
act_vs_pred_1819 <- data.frame(Team_name = laliga1819[,"team_name"],
                               actual_season_points = laliga1819[,"season_points"],
                               predicted_season_points = predict_laliga1819)

act_vs_pred_1718[order(-act_vs_pred_1718$actual_season_points),][1:10,]
act_vs_pred_1819[order(-act_vs_pred_1819$actual_season_points),][1:10,]

```


```{r plot_actual_vs_predicted}
ggplot(act_vs_pred_1718, aes(x = actual_season_points, y = predicted_season_points)) + geom_point() + geom_abline() +
  ggtitle("Actual vs Predicted Season Points for La Liga 2017/2018 Season") +
  geom_text_repel(aes(label = Team_name), size = 3)

ggplot(act_vs_pred_1819, aes(x = actual_season_points, y = predicted_season_points)) + geom_point() + geom_abline() +
  ggtitle("Actual vs Predicted Season Points for La Liga 2018/2019 Season") +
  geom_text_repel(aes(label = Team_name), size = 3)

```

Observation
------------
The MAE of the predicted and actual values for the 2017/2018 & 2018/2019 seasons affirm the effectiveness of the newModel in predicting season points. The scatter plots provide a visual presentation of the predicted and actual values, where teams above the straight line have had higher predicted points than actual points, and the opposite for teams below the line. The model has also predicted higher points for most teams in the 2018/2019 season.

Conclusion/Future Work
----------------------
Regression analysis can be useful in assessing different match statistics and their effects on teams performances. Data analytics is increasingly used by managers in making tactical decisions, and by players during games and training sessions. As a fan, I have an interest in further exploring statistical methods for predicting game outcomes. This can be useful in creating predictor applications for bettors.
